---
title: "Reproducible Research: Peer Assessment 1"
author: "Kalyan Saikia"
keep_md: true
---


## Loading and preprocessing the data

```{r}

library(ggplot2)
library(plyr)
library(knitr)
opts_chunk$set(fig.path = "../Course5_projects/figure/")
```


```{r}
# Reading Data into dataframe
    data_activity<- read.csv("activity.csv")

# Processing data
    data_activity$DateTime<- as.POSIXct(data_activity$date, format="%Y-%m-%d")
```


## What is mean total number of steps taken per day?

```{r}
# Calculating total number of steps taken per day
  Total_step<-aggregate(data_activity$steps ~ data_activity$DateTime, FUN = sum)
      colnames(Total_step)<-c("Date", "Steps")

# Making histogram of the total number of steps taken each day
  hist(Total_step$Steps, breaks = 10, col = "red", xlab = "Steps", main = "Total Steps per Day", ylab = "Frequency")
    
# Calculat the mean and median of the total number of steps taken per day  
  mean(Total_step$Steps, na.rm = TRUE)
  median(Total_step$Steps, na.rm = TRUE)

```
##### From the analysis it is found that Average steps taken per day is 10766 and median value is 10765.

## What is the average daily activity pattern?

```{r}
# Calculating average steps across all data by interval
    ave_step<- aggregate(data_activity$steps ~ data_activity$interval, FUN = mean, na.rm=TRUE)
    colnames(ave_step)<- c("interval", "steps")

# Making plot of average steps
    plot(ave_step$interval, ave_step$steps, type = "l", col="red", xlab = "Interval", ylab = 
    "Average Steps", main = "Average Steps by Interval per day")
    
# Finding out the 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
    ave_step[ave_step$steps==max(ave_step$steps),1]

```

##### The interval 835 contains Maximum Average nuber of steps.


## Imputing missing values
```{r}
# Calculating total number of missing values
    sum(is.na(data_activity$steps))
```

##### In the data set, 2304 NA values are present.  My Strategy to fill the NA values with the average of 5 minutes interval of steps per day.

```{r}

# Creating a copy of the original data
    new_data <- data_activity[,c(1,3,4)]

# Creating a filter with all NA values as True    
    na_filter <- is.na(new_data$steps)

# Calculatine average value of Steps on 5 minute Interval per Day    

    avg_steps_perDay <- tapply(new_data$steps, new_data$interval, mean, na.rm=TRUE, simplify=TRUE)
    
# Creating a NEW DATA SET Replacing NA values with the Mean    
    
    new_data$steps[na_filter] <- avg_steps_perDay[as.character(new_data$interval[na_filter])]
    
# Calculate the total number of steps taken each day
    
    Total_step_new<-aggregate(new_data$steps ~ new_data$DateTime, FUN = sum)
    colnames(Total_step_new)<-c("Date", "Steps")
    
# Making histogram of the total number of steps taken each day using New data
    hist(Total_step_new$Steps, breaks = 10, col = "green", xlab = "Steps", main = "Total Steps per Day", ylab = "Frequency")
  
# Calculate and report the mean and median total number of steps taken per day
    mean(Total_step_new$Steps)
    median(Total_step_new$Steps)
```
##### After replacing NA values, it is noticed that Mean for the new data set remain same whereas the median value is changed (increased).  The effect of NA value imputing is only in the Median in terms of statistical measures. However, no effect is seen on the shape of Histogram.

## Are there any differences in activity patterns between weekdays and weekends?

```{r}
# Converting dates into days using weekdays function
    new_data$days <- weekdays(new_data$DateTime)
    
# Creating a new factor variable indicating whether a given date is a weekday or weekend day.
    new_data$weekDay_Status<-as.factor(ifelse(new_data$days==c("Saturday", "Sunday"), "weekend", "weekday"))

# Creating 5 minutes average of steps for both weekday and weekend
    Avg_steps_weekday_status<-aggregate(new_data["steps"], by=new_data[c("interval","weekDay_Status")], FUN = mean)
   
# Making Timne series plot for Average Steps based on weekday
    library(lattice)
    xyplot(steps ~interval|weekDay_Status, data = Avg_steps_weekday_status, type="l", 
           layout=c(1,2), xlab = "Interval", ylab= "Mean Steps", main="Comparison of Activities on Weekday/Weekend")
```

##### It has been noticed that activity trends are not similar on weekend and Weekday. This may be due to people having an increased opportunity for activity beyond normal work hours for those who work during the week.
